Finally Tagless, Partially Evaluated

### 2008-09-07 tagless/tagless-final.tex
Start to formalize our claims further

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by baking into their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient. We do present the version of our
developement using de Bruijn indices in the accompanying source code, file
|incope-dB.ml|.

@@

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
Fig.~\ref{fig:symantics-haskell}(a)\fi. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\ifshort\else\UseVerbatim{3a}\fi
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as \texttt{app (lam (\textbackslash x -> x)) (bool True)},
whose inferred type is \texttt{Symantics repr => repr Bool}.
For another example, the classical $\mathit{power}$ function is
\ifshort in Fig.~\ref{fig:symantics-haskell}(b)
\else\UseVerbatim{3b}\fi
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
\ifshort in Fig.~\ref{fig:symantics-haskell}(c).
\else\UseVerbatim{3c}\fi
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. \ifshort\else Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag. \fi
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we often show
only one method of each group and elide the rest. The
accompanying code has the complete implementations.

@@

The functor |EX| in Fig.~\ref{fig:ocaml-examples} encodes 
our running examples |test1| and the $\mathit{power}$ function
(|testpowfix|).
The dummy argument to |test1| and |testpowfix| is an artifact of
MetaOCaml, related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction dictates that
the definitions of our object terms must look syntactically like
values. (Alternatively, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.)
\ifshort\else\par\fi
Thus, we represent an object expression in
OCaml as a functor from |Symantics| to an appropriate semantic domain. This
is essentially the same as the constraint \texttt{Symantics repr =>} in the
Haskell embedding.

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} typed, closed object term in the
metalanguage. Moreover, the representation preserves types.
\begin{proposition}\label{prop:typing}
If an object term has object type~$t$, then its
representation in the metalanguage has type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}

\begin{proof}
The rules for deriving a type $t$ from a term $T$ of the object language
correspond \emph{bijectively} to the rules encoded in the signature
|Symantics| (when the latter is read as a set of implications, \`{a} la
Curry-Howard).  In other words, each of the rules from Figure~\ref{fig:object}
corresponds bijectively to one line of |Symantics|.

It is worth noting that, as we use HOAS, rules that normally depend on 
context, in other words rules which are themselves an encoding of a function
from the context to a typing relation, (like those for |lam| and |fix|)
correspond to actual metalanguage functions.
\end{proof}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checker also flags syntactically invalid
object terms, such as if we forget |app| somewhere above.

For simplicity, all examples in the rest of this paper will be in
(Meta)OCaml, but the accompanying code shows a full Haskell implementation
as well.

@@

Now that our term representation is independent of any particular interpreter,
we are now ready to present a series of interpreters.  Each interpreter is an
instance of the |Symantics| class in Haskell and a module implementing
the |Symantics| signature in MetaOCaml.

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation.
\ifshort
\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\else
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

@@

\ifshort\else
A lot of effort has gone into ``typing dynamic typing'': to statically
type-check dynamically\hyp typed values
\citep{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
using the host language's type system to varying extents.
Our object terms are statically typed, so we would
need one of these techniques to interpret dynamically\hyp typed
terms such as those read from a file.
\oleg{First, I suggest we remove the subjunctive mood: `we would
  need...' We have the code already, IncopeTypecheck.hs. We know we
  can typecheck untyped terms so that we do the typechekcing only
  once, and interpret many times. All these interpretation is guranteed to
  be free from any type errors. And this holds even for higher-order
  terms encoded in HOAS. I also suggest we drop
  \cite{haskell-list}. We included the citation solely to show how 
  we could go on. Both links in haskell-list are not actually
  appropriate: first, the object language is \emph{first-order}
  and second they use GADTs. Our IncopeTypecheck uses only
  Typeable and existentials, and no GADTs.
  Finally, we need to find a way to talk about IncopeTypecheck more
  directly, probably in the Introduction -- Sumii-san has directly
  suggested we emphasize that result. For some reason, people (Walid
  and Sumii-san, for ones) consider it important. Incidentally, I
  have the comparison of IncopeTypecheck with `Typing Dynamic
  Typing', which we can use if needed.}
\fi

>>>>

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by expressing in their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, but we have also implemented our
approach using de Bruijn indices (in the accompanying source file
|incope-dB.ml|).

@@

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
Fig.~\ref{fig:symantics-haskell}(a)\fi. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\ifshort\else\UseVerbatim{3a}\fi
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as \texttt{app (lam (\textbackslash x -> x)) (bool True)},
whose inferred type is \texttt{Symantics repr => repr Bool}.
For another example, the classical $\mathit{power}$ function is
\ifshort in Fig.~\ref{fig:symantics-haskell}(b)
\else\UseVerbatim{3b}\fi
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
\ifshort in Fig.~\ref{fig:symantics-haskell}(c).
\else\UseVerbatim{3c}\fi
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. \ifshort\else Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag. \fi
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we often elide all but
one method of each group. The
accompanying code has the complete implementations.

@@

The functor |EX| in Fig.~\ref{fig:ocaml-examples} encodes 
our running examples |test1| and the $\mathit{power}$ function
(|testpowfix|).
The dummy argument to |test1| and |testpowfix| is an artifact of
MetaOCaml: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}),
so we must define our object terms as syntactic values to satisfy
the value restriction.
(Alternatively, we could have used
OCaml's rank-2 record types to maintain the necessary polymorphism.)
\ifshort\else\par\fi
Thus, we represent an object expression in
OCaml as a functor from |Symantics| to a semantic domain. This
is essentially the same as the constraint \texttt{Symantics repr =>} in the
Haskell embedding.

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} well-typed object term in the
metalanguage.
We formalize this representation by defining $H$ and~$M$,
two inductive maps from terms and types in our object language
to terms and types in Haskell and OCaml:
\begin{align}
    H(\ZZ) &= \texttt{Int} &
    M(\ZZ) &= \texttt{int} \notag\\
    H(\BB) &= \texttt{Bool} &
    M(\BB) &= \texttt{bool} \notag\\
    H(t_1 \to t_2) &= H(t_1) \mathbin{\texttt{->}} H(t_2) &
    M(t_1 \to t_2) &= M(t_1) \mathbin{\texttt{->}} M(t_2) \displaybreak[0] \\
    H(x) &= x &
    M(x) &= x \notag\\
    H(\fun{x}e) &= \texttt{lam (\textbackslash x -> $H(e)$)} &
    M(\fun{x}e) &= \texttt{lam (fun x -> $M(e)$)} \notag\\
    H(\fix{f}e) &= \texttt{fix (\textbackslash x -> $H(e)$)} &
    M(\fix{x}e) &= \texttt{fix (fun x -> $M(e)$)} \notag\\
    H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
    M(e_1 e_2) &= \texttt{app $M(e_1)$ $M(e_2)$} \notag\\
    H(n) &= \texttt{int $n$} &
    M(n) &= \texttt{int $n$} \notag\\
    H(\True) &= \texttt{bool True} &
    M(\True) &= \texttt{bool true} \notag\\
    H(\False) &= \texttt{bool False} &
    M(\False) &= \texttt{bool false} \notag\\
    \!H(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $H(e)$ $H(e_1)$ $H(e_2)$}} \notag\\
    \!M(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $M(e)$ (fun () -> $M(e_1)$) (fun () -> $M(e_2)$)}} \notag\\
    H(e_1 + e_2) &= \texttt{add $H(e_1)$ $H(e_2)$} &
    M(e_1 + e_2) &= \texttt{add $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \times e_2) &= \texttt{mul $H(e_1)$ $H(e_2)$} &
    M(e_1 \times e_2) &= \texttt{mul $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \le e_2) &= \texttt{leq $H(e_1)$ $H(e_2)$} &
    M(e_1 \le e_2) &= \texttt{leq $M(e_1)$ $M(e_2)$}
\end{align}
These definitions assume that our object language, Haskell, and OCaml
use the same variable names~$x$ and integer literals~$n$.
If $\Gamma$ is a typing context $x_1:t_1,\dotsc,x_n:t_n$ in the object
language, then we define the metalanguage contexts
\begin{align}
    \texttt{repr\;$H(\Gamma)$}
    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \notag\\
    \texttt{('c,\:$M(\Gamma)$)\;repr}
    &= x_1:\texttt{('c,\:$M(t_1)$)\;repr},\dotsc,x_n:\texttt{('c,\:$M(t_n)$)\;repr} \text.
\end{align}

The following proposition states the trivial but fundamental fact that
this representation preserves types.
\begin{proposition}\label{prop:typing}
If an object term~$e$ has the type~$t$ in the context~$\Gamma$,
then the Haskell term~$H(e)$ has the type \texttt{repr\;$H(t)$}
in the context
\[
    \texttt{repr}:\star\to\star,\quad
    \texttt{Symantics repr},\quad
    \texttt{repr}\;H(\Gamma)
    \text,
\]
and the OCaml term~$M(e)$ has the type \texttt{('c,\:$M(t)$)\;repr}
in the context
\[
    \texttt{S:Symantics},\quad
    \texttt{open S},\quad
    \texttt{'c}:\star,\quad
    \texttt{('c,\:$M(\Gamma)$)\;repr}
    \text.
\]
\end{proposition}
\begin{proof}
By structural induction on the derivation in the object language
that $e$ has type~$t$ in~$\Gamma$.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has the type~$t$,
then the Haskell term~$H(e)$ has the type
\[
    \texttt{forall repr. Symantics repr => repr $H(t)$}
\]
and the OCaml functor
\[
    \texttt{functor (S:Symantics) -> struct open S let term () = $M(e)$ end}
\]
has the signature
\[
    \texttt{functor (S:Symantics) -> sig val term: unit -> ('c,\:$M(t)$)\;S.repr end}
    \text{.}
\]
\end{corollary}

Conversely, the type system of the metalanguage checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checker also flags syntactically invalid
object terms, such as if we forget |app| somewhere above.

For simplicity, all examples in the rest of this paper are in
(Meta)OCaml, but the accompanying code shows a full Haskell implementation
as well.

@@

Now that our term representation is independent of any particular interpreter,
we are ready to present a series of interpreters.  Each interpreter is an
instance of the |Symantics| class in Haskell and a module implementing
the |Symantics| signature in OCaml.

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below is metacircular in that it
interprets each object\hyp language
operation as the corresponding metalanguage operation.
\ifshort
\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\else
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

@@
